Leadership across cultures and contexts: an exploratory analysis

### 2020-05-03 leadership_across_cultures_contexts.Rmd
Reorganized intro sections on functions.

There is now well established consensus within the behavioral sciences on the importance of data from the full range of human cultural diversity. Systemic biases in much of social science are well known, and might be due, in part, to the lack of data from populations other than American undergraduates [@henrich_weirdest_2010, *The weirdest people in the world?*].

Since the seminal WEIRD people paper, many social scientists have taken up the challenge to diversify their samples and have sought out "non-WEIRD" participants and databases. Dichotomizing populations and their psychologies and social dynamics as WEIRD or non-WEIRD, however, is a mistake.^[A Google Scholar search for "non-WEIRD" produced over 1000 results.] As anthropology has conclusively demonstrated, there is enormous diversity across societies that does not remotely resemble a dichotomy.

@@

There is potentially as much diversity within nonindustrial, non-WEIRD cultures as there is within WEIRD ones, and there is also substantial overlap of the behaviors of people everywhere [@brown_human_1991]. @henrich_weirdest_2010 in their critique of the over-reliance on WEIRD samples emphasized a holistic approach to cultural diversity within and across "telescoping" levels of human societies. It is therefore important to assess cultural variation without assuming that everyone within a culture is essentially the same, and without putting undue emphasis on either the between-population differences or similarities.

@@

<!-- CLOSING -->
We first briefly review theoretical, ethnographic, and empirical perspectives on the qualities of leaders, the functions they serve, and the costs and benefits of leadership across human societies, including WEIRD and non-WEIRD populations. We then outline the aims and methods of the current cross-cultural study before presenting results and interpretations. Through novel exploratory analyses encompassing the full range of cultural diversity and diverse social contexts we aim to advance leadership studies across disciplines and provide researchers with a foundation for more detailed approaches to human leadership across cultures and contexts.

<!-- Reference a historical example of a short term leader implementing change with lasting effects, good and/or bad. -->

## The functions of leadership across cultures and contexts
<!-- INTRO -->
Despite the universality of the phenomenon of leadership, the function leaders serve across populations and social contexts is highly diverse and are expected to be closely tied to group structure and group goals [@van_vugt_cognitive_2007]. Among many of the ethnographically described small-scale societies, which generally lack multi-level, institutionalized social and political structures, leadership at the community level is typically informal and ephemeral and leaders emerge contextually when demands of coordination or conflict resolution require a managerial presence [@von_rueden_leadership_2015].

<!-- FUNCTIONS in SSS = CORDINATION AND CONFCLIT RESOLUTION -->
In the absence of institutionalized organizational structure leaders often take on essential roles in group coordination and collective action, including facilitating conflict resolution within or between autonomous households or kin groups, managing individual investments to group-wide projects, and administering sanctions to individuals who do not appropriately contribute to collective actions, comply with group policies, or seriously violate social norms. @glowacki_leadership_2015 draw on empirical, ethnographic, and theoretical models of leader emergence and suggest in small-scale societies institutionalized leadership roles are likely to initially emerge to resolve conflicts between families. @roscoe_social_2009 draws on the ethnography of highland New Guinea in describing the modular structure of human societies and suggests the functions of community leaders in small-scale societies are structured by the joint demands of signaling military formidability to rival groups and facilitating within-group conflict resolution and cooperation. For "big-men" in these societies, in addition to military prowess, "talents for oratory, mediation, and organization were instrumental in reducing their internal conflict and enhancing the ability of their members to coordinate their actions as a group" [@roscoe_social_2009, p. 104].

<!-- GETS INTO QUALITIES BELOW -->
Given demands of community organization and conflict resolution, leadership tends to be context-specific and leaders in such contexts often draw on persuasive capacities and leverage community-based respect, specialized skills or knowledge, and social support to achieve or maintain community influence [@kantner_identifying_2010; @lewis_leaders_1974; @boehm_egalitarian_1993].

<!-- REPETITIVE COMPARED TO ABOVE BUT INCLUDES ADDITIONAL DETAILS about FUNCTIONS AND  MORE CITATIONS, THEORY -->
The theoretical and ethnographic literature suggests these demands most commonly arise in the contexts of collective actions and in conflict resolution [e.g., @glowacki_leadership_2015; @gavrilets_convergence_2016; @hooper_theory_2010; @von_rueden_solving_2015; @henrich_big_2015; @pietraszewski_evolution_2019]. Demands of mediation in within-group conflict resolution, although variable in scale, are probably a universal feature of human groups [@king_origins_2009; @van_vugt_leadership_2008]. Demands of between-group interactions, including both conflict and cooperation are also very common across small-scale societies, including hunter-gatherers [@glowacki_evolutionary_2017; @hames_pacifying_2019; @richerson_cultural_2016; @henrich_evolution_2009; @apicella_social_2012]. Across relatively small-scale societies representing the group and managing both internal and external relationships and conflicts are expected to be the most salient functions of community leaders. Within communities, however, leaders may emerge whenever individuals congregate. The role of leaders in ritual contexts, such as in initiations [@singh_cultural_2017], and in providing social services, including facilitating marriages [@walker_evolutionary_2011] and organizing social activities [@wiessner_food_1996], are also expected to be common functions carried out by a leader.

@@

## Culture, context, and human leadership

@@

Across cultures and contexts, when ethnographic texts described female leaders exclusively, they were relatively more likely to be described as older, of high status, married to a high-quality spouse in a polygynous marriage, and the recipients of various mating benefits. This suite of characteristics is consistent with discussions of high status first-wives who gain and maintain increased social influence across the lifespan by leveraging their extended kin and social networks [@goodale_tiwi_1971; @brown_her_1985], and supports the view of positive assortative mating of high status leaders @garfield_investigating_2019 documented among elected leaders in a small-scale society with relative gender-egalitarianism. Evidence for male leaders (relative to evidence for female leaders or evidence that was not sex-specific) was indeed linked to group-level and political functions and male leaders were relativey more likely to provide social services, control political structure, and were at greater risk of losing their high social standing. See Figure S\@ref(fig:heatmapsex).

Female leaders were also more likely to lack coercive authority, which is consistent with much of leadership studies suggesting dominant and aggressive women are less likely to be preferred as leaders, unlike dominent men who are sometimes prefered as leaders [@garfield_investigating_2019; @williams_subtle_2016].

Despite sex differences, there is also likely to be substantial overlap in many of the qualities and functions of female and male leaders across human societies. Leader sex was not a predictor of the majority of the measures investigated here (adjusting for other population-level measures; see Table S\@ref(tab:universalvars)). When sex was a predictor (Table S\@ref(tab:variablevars)), descriptions including both female and male leaders provided relatively more supporting evidence for many key measures of leader qualities and functions, including group organization, resource distribution, ritual functions, and being brave and experienced and accomplished.

>>>>

There is now well established consensus within the behavioral sciences on the importance of data from the full range of human cultural diversity. Systemic biases in much of social science are well known, and might be due, in part, to the lack of data from populations other than American undergraduates [@henrich_weirdest_2010, *The weirdest people in the world?*]. Since the seminal WEIRD people paper, many social scientists have taken up the challenge to diversify their samples and have sought out "non-WEIRD" participants and databases. Dichotomizing populations and their psychologies and social dynamics as WEIRD or non-WEIRD, however, is a mistake.^[A Google Scholar search for "non-WEIRD" produced over 1000 results.] As anthropology has conclusively demonstrated, there is enormous diversity across societies that does not remotely resemble a dichotomy.

@@

There is potentially as much diversity within nonindustrial, non-WEIRD populations as there is within WEIRD ones, and there is also substantial overlap of the behaviors of people everywhere [@brown_human_1991]. @henrich_weirdest_2010 in their critique of the over-reliance on WEIRD samples emphasized a holistic approach to cultural diversity within and across "telescoping" levels of human societies. It is therefore important to assess cultural variation without assuming that everyone within a culture is essentially the same, and without putting undue emphasis on either the between-population differences or similarities.

@@

<!-- CLOSING -->
We first briefly review theoretical, ethnographic, and empirical perspectives on the qualities of leaders, the functions they serve, and the costs and benefits of leadership across human societies, including WEIRD and non-WEIRD populations. We then outline the aims and methods of the current cross-cultural study -- which focuses on uncovering patterns among non-WEIRD populations -- before presenting results and interpretations. Through novel exploratory analyses encompassing the full range of cultural diversity and diverse social contexts we aim to advance leadership studies across disciplines and provide researchers with a foundation for more detailed approaches to human leadership across cultures and contexts.

## The functions of leadership across cultures and contexts
<!-- INTRO -->
Despite the universality of the phenomenon of leadership, the function leaders serve across populations and social contexts is highly diverse and are expected to be closely tied to group structure and group goals [@van_vugt_cognitive_2007]. Among many of the ethnographically described small-scale societies, which generally lack multi-level, institutionalized social and political structures, leadership at the community level is typically informal and ephemeral and leaders emerge contextually when demands of coordination or conflict resolution require a managerial presence [@von_rueden_leadership_2015; @kantner_identifying_2010; @lewis_leaders_1974; @boehm_egalitarian_1993].

The theoretical and ethnographic literature suggests these demands most commonly arise in the contexts of collective actions and in conflict resolution [e.g., @glowacki_leadership_2015; @gavrilets_convergence_2016; @hooper_theory_2010; @von_rueden_solving_2015; @henrich_big_2015; @pietraszewski_evolution_2019]. Demands of mediation in within-group conflict resolution, although variable in scale, are probably a universal feature of human groups [@king_origins_2009; @van_vugt_leadership_2008]. In the absence of institutionalized organizational structure leaders often take on essential roles in within-group coordination and collective action, including facilitating conflict resolution within or between autonomous households or kin groups, managing individual investments to group-wide projects, and administering sanctions to individuals who fail to appropriately contribute to collective actions, comply with group policies, or seriously violate social norms. Demands of between-group interactions, including both conflict and cooperation are also very common across small-scale societies, including hunter-gatherers [@glowacki_evolutionary_2017; @hames_pacifying_2019; @richerson_cultural_2016; @henrich_evolution_2009; @apicella_social_2012]. Across relatively small-scale societies representing the group and managing both internal and external relationships and conflicts are expected to be the most salient functions of community leaders. Within communities, however, leaders may emerge whenever individuals congregate. The role of leaders in ritual contexts, such as in initiations [@singh_cultural_2017], and in providing social services, including facilitating marriages [@walker_evolutionary_2011] and organizing social activities [@wiessner_food_1996], are also expected to be common functions carried out by a leader.

@glowacki_leadership_2015 draw on empirical, ethnographic, and theoretical models of leader emergence and suggest in small-scale societies institutionalized leadership roles are likely to initially emerge to resolve conflicts between families. @roscoe_social_2009 draws on the ethnography of highland New Guinea in describing the modular structure of human societies and suggests the functions of community leaders in small-scale societies are structured by the joint demands of signaling military formidability to rival groups and facilitating within-group conflict resolution and cooperation. For "big-men" in these societies, in addition to military prowess, "talents for oratory, mediation, and organization were instrumental in reducing their internal conflict and enhancing the ability of their members to coordinate their actions as a group" [@roscoe_social_2009, p. 104].

<!-- GETS INTO QUALITIES BELOW -->
<!-- Given demands of community organization and conflict resolution, leadership tends to be context-specific and leaders in such contexts often draw on persuasive capacities and leverage community-based respect, specialized skills or knowledge, and social support to achieve or maintain community influence [@kantner_identifying_2010; @lewis_leaders_1974; @boehm_egalitarian_1993].  -->

@@

## Subsistence strategy and group context effects on human leadership

@@

Across cultures and contexts, when ethnographic texts described female leaders exclusively, they were relatively more likely to be described as older, of high status, married to a high-quality spouse in a polygynous marriage, and the recipients of various mating benefits. This suite of characteristics is consistent with discussions of high status first-wives who gain and maintain increased social influence across the lifespan by leveraging their extended kin and social networks [@goodale_tiwi_1971; @brown_her_1985], and supports the view of positive assortative mating of high status leaders @garfield_investigating_2019 documented among elected leaders in a small-scale society with relative gender-egalitarianism. Evidence for male leaders (relative to evidence for female leaders or evidence that was not sex-specific) was indeed linked to group-level and political functions and male leaders were relativey more likely to provide social services, control political structure, and were at greater risk of losing their high social standing. Female leaders, relative to males, were also more likely to lack coercive authority, which is consistent with much of leadership studies suggesting dominant and aggressive women are less likely to be preferred as leaders, unlike dominent men who are sometimes prefered as leaders [@garfield_investigating_2019; @williams_subtle_2016]. See Figure S\@ref(fig:heatmapsex).

Despite sex differences, there is also likely to be substantial overlap in many of the qualities and functions of female and male leaders across human societies. Leader sex was not a predictor of the majority of the measures investigated here (adjusting for other population-level measures; see Table S\@ref(tab:universalvars)). When sex was a predictor (Table S\@ref(tab:variablevars)), descriptions including both female and male leaders provided relatively more supporting evidence for many key measures of leader qualities and functions, including group organization, resource distribution, ritual functions, and being brave and experienced and accomplished )Figure S\@ref(fig:heatmapsex).
